---
title: "Tutorial of UltraGBIF"
author:
  - name: anonymous
    affiliation: anonymous University
    email: anonymous@anonymous.com
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tutorial of UltraGBIF}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
---

## Description

This Tutorial parses and merges *Saxifraga* (<https://doi.org/10.15468/dl.bythb4>) occurrence records from GBIF.

## import records

This tutorial uses *Saxifraga* occurrence records download from GBIF that has been integrated in the package.

This module also utilizes `EnumOccurrence` Issue values to transform these issues into a structured format, enabling systematic quality assessment through individualized issue tracking and standardized classification. See more [details](https://gbif.github.io/gbif-api/apidocs/org/gbif/api/vocabulary/OccurrenceIssue.html) here.

```{r import_records, eval=FALSE, include=T}
library(UltraGBIF)
library(curl)
url <- 'https://api.gbif.org/v1/occurrence/download/request/0021523-250402121839773.zip'
gbif_occurrence_file <- paste0(getwd(),'/Saxfriga.zip')
curl_download(url,gbif_occurrence_file,quiet = F) ## cost time and total are 68692523 bytes
occ_import <- import_records(GBIF_file = gbif_occurrence_file,
                             only_PRESERVED_SPECIMEN = T)
```

`occ_import` includes `occ` for initial occurrence records, `occ_gbif_issue` for GBIF issues and its `summary`, and duration.

## Check taxon name

UltraGBIF uses TNRS[@boyle2013] to check and resolve taxon names in occurrence records. The Taxonomic Name Resolution Service (TNRS) is a tool for the computer-assisted standardization of plant scientific names. The TNRS corrects spelling errors and alternative spellings to a standard list of names, and converts out of date names (synonyms) to the current accepted name. The TNRS can process many names at once, saving hours of tedious and error-prone manual name correction.

```{r check_occ_name, eval=FALSE, include=T}
taxa_checked <- check_occ_name(occ_import = occ_import,accuracy = 0.9)
```

`taxa_checked` includes `occ_wcvp_check_name` for records with resolved taxa and its `summary`, and duration.

## Check collectors dictionary

This module extracts the primary collector's surname from the `recordedBy` field in occurrence records and generates an association list linking standardized surnames to corresponding records. It assists to resolve collector identity ambiguities and enforce nomenclatural consistency. Referred `CollectorDictionary` in this module comes from ParseGBIF[@demelo2024].

```{r prepare_collectors_dictionary, eval=FALSE, include=T}
dictionary <- prepare_collectors_dictionary(occ_import = occ_import,
                                            surname_selection_type = "largest_string",
                                            max_words_name = 6,
                                            min_characters_in_name = 2)
```

`dictionary` includes `my_dictionary` for checked collectors name and `ref_dictionary` for referred dictionary, and duration.

## Generate unique collection mark

To resolve duplicate records in biodiversity datasets, UltraGBIF generates a standardized composite key for each unique collection event like ParseGBIF[@demelo2024] does. This key is constructed by concatenating three core metadata fields:

-   **Plant family** (`family`)

-   **Standardized collector surname** (`recordedByStandardized`)

-   **Numeric-only collection number** (`recordNumber_Standard`)

The resulting identifier, `key_family_recordedBy_recordNumber`, groups specimens sharing identical collection events. An optional temporal variant, `key_year_recordedBy_recordNumber`, incorporates collection year where available to enhance resolution.

Standardization ensures robust matching: non-numeric characters are stripped from collection numbers, and collector names are normalized using predefined dictionaries to reconcile spelling variants.

This module addresses two key challenges:

1.  Disambiguating common collector surnames through taxonomic context.

2.  Enabling automated duplicate detection without reliance on often-missing temporal metadata.

```{r generate_collection_mark, eval=FALSE, include=T}
collection_key <- generate_collection_mark(occ_import = occ_import,
                                            dictionary = dictionary)
```

`collection_key` includes `occ_collectorsDictionary` for input records updated with standardized fields and generated keys and `CollectorsDictionary_add` for newly encountered collectors for dictionary expansion and duration.

## Set digital voucher

Following generation of a unique collection event identifier (collection_key), this module resolves duplicate records and outlines the selection of the master digital voucher and associated integration to represent each event.

**Duplicate Grouping and Voucher Selection**

-   For records possessing complete collection event keys, duplicates are identified and grouped based on identical `collection_key` values. Within each group, a master digital voucher is selected by evaluating a composite score that combines two dimensions: metadata completeness and geospatial reliability. Metadata completeness is quantified as the sum of binary presence indicators across 10 fields, including `recordedBy`, `locality`, and `catalogNumber`, where presence is scored as 1 and absence as 0. Geospatial reliability is assigned according to GBIF’s `EnumOccurrenceIssue` classification, with values of −1 indicating unaffected coordinates, −3 indicating potential positional error, and −9 indicating coordinates unsuitable for spatial analysis. The record achieving the highest combined score is designated the master voucher. Any missing metadata in the master record are supplemented using non-redundant values drawn from other records within the same group.

-   For records with incomplete keys (e.g., missing `recordedBy` or `recordNumber`), no grouping is performed, and each is treated as a singleton event. Nevertheless, metadata completeness and geospatial scores are computed identically to ensure consistent, system-wide data quality assessment.

**Taxonomic Assignment**

-   For groupable events, the accepted taxon name (`UltraGBIF_sample_taxon_name`) is assigned as the most frequently occurring species- or subspecies-level identification among duplicates; ties are resolved alphabetically. Unidentified events are flagged (`UltraGBIF_unidentified_sample = TRUE`).

-   Non-groupable records are processed individually with identical taxonomic logic.

**Geospatial Integration**

If the master voucher has unusable coordinates (score = -9) or lacks coordinates entirely, usable coordinates from duplicates (score ≥ -3) are preferentially retained.

**Standardized Output Fields**

Key output variables include:

-   `UltraGBIF_digital_voucher`: identifies master record (logical);

-   `UltraGBIF_duplicates`: presence of duplicates (logical);

-   `UltraGBIF_num_duplicates`: count of grouped records;

-   `UltraGBIF_non_groupable_duplicates`: records with incomplete keys (logical);

-   `UltraGBIF_duplicates_grouping_status`: grouping feasibility (e.e., "groupable", "not groupable: no recordedBy");

-   `UltraGBIF_sample_taxon_name_status`: taxonomic resolution ("identified", "divergent", "not identified");

-   `UltraGBIF_useful_for_spatial_analysis`: coordinate usability (logical).

This module ensures robust duplicate resolution, metadata enrichment, and interoperability with global biodiversity standards.

```{r set_digital_voucher, eval=FALSE, include=T}
voucher <- set_digital_voucher(occ_import = occ_import,
                               taxa_checked = taxa_checked,
                               collection_key = collection_key)
```

`voucher` includes the full flagged digital voucher of occurrence records which names `occ_digital_voucher` and its `summary`, and duration.

## Refine records

When a `voucher` is flagged as usable, its missing information in fields is restored by consolidating data from duplicate records belonging to identical collection events, creating one clean record per unique event, whether the event key is complete or not. It picks the most common species-level name (using alphabetical order to break ties) or labels the event “unidentified” if no such name exists; incomplete keys are treated as separate events. Missing coordinates in the main record are filled in from duplicates.

For finally complete usable vouchers, their associated occurrence records are automatically subjected to geospatial validation and precise WGSRPD attribution. Leveraging the R packages CoordinateCleaner[@zizka2019] and terra, this module efficiently performs coordinate validation and cleaning, targeting common geospatial anomalies such as misassigned capitals, geographic centroids, duplicate coordinates, GBIF headquarters artifacts, institution-proximate records, distributional outliers, marine points in terrestrial datasets, and zero/placeholder coordinates.

Then each validated record is matched to its corresponding WGSRPD level-3 Botanical Country classifications with different status (native/introduced/extinct/location_doubtful), which minimizes mis-classification risks in cross-regional analyses.

To optimize computational efficiency, this module supports parallelization through multi-threading by specifying the desired concurrency level via the `threads` parameter.

```{r refine_records, eval=FALSE, include=T}
refine <- refine_records(voucher = voucher,
                            threads = 3,
                            save_path = dirname (gbif_occurrence_file),
                            tests = c("capitals", "centroids", "equal", "gbif", "institutions", "outliers", "seas","zeros"))
```

`refine` includes `all_records` for all cleaned records and `native_records` for native of them, and time costs. If you set `save_path`, then `all_records` and `native_records` are exported, named as `usable_refined_records.csv.gz` and `native_refined_records.csv.gz`, both are potentially meritorious for your future research.

## Map records

This module is implemented optionally, rendering the output from `coordinate_refine`, employing `geohashTools`[@geohashTools] to perform spatial clustering of geo-coordinates. The `precision` parameter determines grid resolution (e.g. 4 for 20 km×20 km grids, 3 for 156 km×156 km grids). It groups records by `UltraGBIF_wcvp_taxon_name`, WGSRPD information and `geohash grid` and follows deduplication.

After simplification, dynamic maps with different layers are rendered by mapview[@mapview] with WGSRPD information legend. Symbol size is controlled by the `cex` parameter. Moreover, they supports zoom functionality for spatial resolution adjustment and enables interactive selection of individual coordinates to display their full attributes.

```{r map_records, eval=FALSE, include=T}
map_records(records_refined = refine,
                     precision = 4, 
                     cex = 4)
```

## Map richness

This optional module is useful for creating a simple richness map from your processed occurrence records above. It is based on two functions `lets.presab.points` and `plot.PresenceAbsence` from R package `letsR`[@vilela2015], but fully leverages vectorization techniques to avoid looping when filling large matrices, thus achieving nearly a hundredfold speedup.

```{r plot_richness, eval=FALSE, include=T}
richness <- plot_richness(records_refined = refine,main = 'Saxifraga Richness')
```

A richness plot will be shown after running this module, and additionally, `richness` includes `presence` for the taxa\~grid presence table and `ras_richness` for a richness S4 object.

## Reference
